Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-05-04 doc/paper.Rmd
line breaks at sentences

Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992]. These effects can be complex or even contradictory - mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@condit2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated (see also @connell2000; @engelbrecht2002; @edwards2006]. There is also evidence that the effects can persist for multiple years [ @phillips2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].

@@

Third, it may be that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on other species rather than the direct effects on individual physiology [@evers2021]. For example, while droughts can kill some tropical trees immediately [@laurance2001], others take three years or more [@crileyYearProductionHigh1994]. This will potentially lead to a boom in growth and flowering years after the drought, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990]. Similar delayed changes in the local environment could also influence the foraging behavior of *H. acuminata*'s pollinators [@bruna2004; @stouffer1996] and seed dispersers [@uriarte2011]. While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, [**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?\*\*]\*\* it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.

>>>>

Precipitation extremes and other climatic factors are known to have immediate effects on the growth, survival, or reproduction of plants [@wrightSeasonalNinoLonger2006; @estebanOtherSideDroughts2021], including *Heliconia* [@stiles1975; @westerbandEarlyLifeConditions2017] and other tropical herbs [@wrightSeasonalDroughtSoil1992].
These effects can be complex or even contradictory - mild droughts can increase the growth rates of tropical trees, perhaps due to reductions in cloud cover [@condit2004], but in severe drought years growth can be extremely low and mortality can be sharply elevated (see also @connell2000; @engelbrecht2002; @edwards2006]. There is also evidence that the effects can persist for multiple years [@phillips2010], such as a boom in drought-year fruit production followed by severe post-drought "famine" @wright1999; see also @pau2013].

@@

Third, it may be that the delayed demographic effects we observed are indirectly mediated by the effect of SPEI on other species rather than the direct effects on individual physiology [@evers2021].
For example, while droughts can kill some tropical trees immediately [@laurance2001], others take three years or more [@crileyYearProductionHigh1994].
This will potentially lead to a boom in growth and flowering years after the drought, when the canopy gaps resulting from delayed leaf drop [@janssenDroughtEffectsLeaf2021] and treefalls allow for light penetration to the forest understory [@canham1990].
Similar delayed changes in the local environment could also influence the foraging behavior of *H. acuminata*'s pollinators [@bruna2004; @stouffer1996] and seed dispersers [@uriarte2011].
While more work is needed to explain why the delayed effects of SPEI on survival and growth are greater in forest fragments than interiors, [**question about wording** - does this mean the delay is relatively greater, or the magnitude of the effect is greater?\*
\*]\*\* it may be due to the proportionately greater amount of litterfall on edges [@vasconcelos2004] altering the abundance of pathogens or mycorrhizae.
